---
title: "Aequity-evaluation-mammo-ai"
author: "Hyemi"
date: "2026-02-05"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# Fix: set CRAN mirror so install.packages() works during knitting
options(repos = c(CRAN = "https://cloud.r-project.org"))

pkgs <- c(
  "dplyr","readr","knitr","pROC","ggplot2","scales",
  "broom","tidyr","forcats"
)

need <- setdiff(pkgs, rownames(installed.packages()))
if(length(need) > 0) install.packages(need)
invisible(lapply(pkgs, library, character.only = TRUE))
```

# 1 Background and objectives

## 1.1 Background

This project evaluates whether a mammography AI algorithm’s cancer detection score performs **consistently across demographic subgroups**.
Using a simulated screening dataset (N = 5,000), we quantify:

- **Discrimination:** how well the AI score separates cancer vs non-cancer (AUC-ROC).
- **Calibration:** whether the score behaves like a probability (predicted risk matches observed cancer rates).
- **Potential structure/confounding:** whether region- and vendor-related differences may influence apparent subgroup differences.

*Note:* In many real systems, a model output may be a ranking/suspicion score rather than a calibrated probability. In that case, high AUC can coexist with imperfect calibration.

## 1.2 Objectives

- Estimate overall and subgroup AUC-ROC (ethnicity, SES, and ethnicity × SES).
- Compare subgroup AUCs to reference groups (Swedish; High SES) using DeLong tests.
- Assess overall and subgroup calibration using calibration curves, Brier score, and Expected Calibration Error (ECE).
- Examine whether associations persist after adjustment for region/vendor/age/density.

# 2 Data & preprocessing

```{r data_processing}
# GitHub-friendly: relative paths (repo root)
data_path <- file.path("data", "mammography_ai_equity_data_simulated.csv")
stopifnot(file.exists(data_path))

data_original <- readr::read_csv(data_path, show_col_types = FALSE)

# Type casting + derived intersectional subgroup
data_original <- data_original %>%
  mutate(
    cancer = as.integer(cancer),
    ethnicity = as.factor(ethnicity),
    ses = factor(ses, levels = c("Low","Medium","High")),
    region = as.factor(region),
    vendor = as.factor(mammography_vendor),     # anonymized A/B/C/D
    density = factor(breast_density, levels = c("A","B","C","D")),
    eth_ses = interaction(ethnicity, ses, drop = TRUE, sep = "/")
  ) %>%
  select(-mammography_vendor, -breast_density)

knitr::kable(
  data_original %>% summarise(across(everything(), ~sum(is.na(.)))),
  caption = "Missing values per variable"
)
```

Preprocessing summary:

- Cast categorical variables as factors and ensured `cancer` is a 0/1 integer.
- Created `eth_ses` (ethnicity × SES) for intersectional summaries.
- Confirmed no missingness in the analytic variables.

## 2.1 Subgroup sizes

```{r subgroup_sizes}
by_eth <- data_original %>%
  group_by(ethnicity) %>%
  summarise(n=n(), cases=sum(cancer), prev=mean(cancer), .groups="drop")

by_ses <- data_original %>%
  group_by(ses) %>%
  summarise(n=n(), cases=sum(cancer), prev=mean(cancer), .groups="drop")

by_int <- data_original %>%
  group_by(eth_ses) %>%
  summarise(n=n(), cases=sum(cancer), prev=mean(cancer), .groups="drop")

kable(by_eth, digits=3, caption="Group size by ethnicity")
kable(by_ses, digits=3, caption="Group size by SES")
kable(by_int, digits=3, caption="Group size by ethnicity × SES")
```

# 3 Discrimination (AUC-ROC)

## 3.1 Overall AUC

```{r auc_overall}
roc_all <- roc(response = data_original$cancer, predictor = data_original$ai_score, quiet=TRUE)
auc_all <- auc(roc_all)
ci_all <- ci.auc(roc_all, method="delong")

tibble(
  auc = as.numeric(auc_all),
  ci_low = as.numeric(ci_all)[1],
  ci_high = as.numeric(ci_all)[3]
) %>%
  kable(digits=3, caption="Overall AUC (DeLong 95% CI)")

roc_df <- tibble(
  fpr = 1 - roc_all$specificities,
  tpr = roc_all$sensitivities
)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line(linewidth=1) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  labs(
    title = "Overall ROC curve",
    subtitle = paste0(
      "AUC = ", round(as.numeric(auc_all), 3),
      " (95% CI ", round(as.numeric(ci_all)[1],3), "–", round(as.numeric(ci_all)[3],3), ")"
    ),
    x = "False positive rate (1 - specificity)",
    y = "True positive rate (sensitivity)"
  ) +
  theme_classic(base_size = 12)
```

Interpretation: Discrimination is very high in this simulated dataset (AUC close to 1), meaning the score ranks cancer cases above non-cases very consistently.

## 3.2 Subgroup AUCs

```{r auc_subgroups}
get_auc_ci <- function(df){
  r <- roc(df$cancer, df$ai_score, quiet=TRUE)
  c(
    auc = as.numeric(auc(r)),
    ci_low = as.numeric(ci.auc(r, method="delong"))[1],
    ci_high = as.numeric(ci.auc(r, method="delong"))[3]
  )
}

auc_ethnicity <- data_original %>%
  group_by(ethnicity) %>%
  group_modify(~{
    m <- get_auc_ci(.x)
    tibble(
      n = nrow(.x),
      cases = sum(.x$cancer),
      auc = m["auc"],
      ci_low = m["ci_low"],
      ci_high = m["ci_high"]
    )
  }) %>%
  ungroup()

auc_ses <- data_original %>%
  group_by(ses) %>%
  group_modify(~{
    m <- get_auc_ci(.x)
    tibble(
      n = nrow(.x),
      cases = sum(.x$cancer),
      auc = m["auc"],
      ci_low = m["ci_low"],
      ci_high = m["ci_high"]
    )
  }) %>%
  ungroup()

min_cases <- 20
auc_int <- data_original %>%
  group_by(ethnicity, ses) %>%
  filter(sum(cancer) >= min_cases, sum(1-cancer) >= min_cases) %>%
  group_modify(~{
    m <- get_auc_ci(.x)
    tibble(
      n = nrow(.x),
      cases = sum(.x$cancer),
      auc = m["auc"],
      ci_low = m["ci_low"],
      ci_high = m["ci_high"]
    )
  }) %>%
  ungroup()

kable(auc_ethnicity, digits=3, caption="AUC by ethnicity (DeLong 95% CI)")
kable(auc_ses, digits=3, caption="AUC by SES (DeLong 95% CI)")
kable(auc_int, digits=3,
      caption=paste0("AUC by ethnicity × SES (cells with ≥", min_cases, " cases and non-cases)") )

ggplot(auc_ethnicity %>% mutate(ethnicity = reorder(ethnicity, auc)),
       aes(x = auc, y = ethnicity)) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.2) +
  geom_point(size = 2.3) +
  geom_vline(xintercept = as.numeric(auc_all), linetype="dashed") +
  coord_cartesian(xlim = c(0.9, 1)) +
  labs(
    title = "Discrimination by ethnicity",
    subtitle = "Points: AUC; bars: 95% CI (DeLong). Dashed line: overall AUC.",
    x = "AUC-ROC",
    y = NULL
  ) +
  theme_classic(base_size = 12)
```

## 3.3 AUC comparisons vs reference groups

```{r delong_tests}
# Ethnicity vs Swedish
sw <- subset(data_original, ethnicity == "Swedish")
roc_sw <- roc(sw$cancer, sw$ai_score, quiet = TRUE)

groups_eth <- setdiff(levels(data_original$ethnicity), "Swedish")
res_delong <- tibble(ethnicity = character(), p_value = numeric())

for(gname in groups_eth){
  gdat <- subset(data_original, ethnicity == gname)
  if(length(unique(gdat$cancer)) < 2) next
  roc_g <- roc(gdat$cancer, gdat$ai_score, quiet = TRUE)
  p <- roc.test(roc_g, roc_sw, method = "delong", paired = FALSE)$p.value
  res_delong <- bind_rows(res_delong, tibble(ethnicity = gname, p_value = p))
}

res_delong <- res_delong %>% mutate(p_adj = p.adjust(p_value, method = "BH"))

kable(res_delong, digits = 4,
      caption = "DeLong test p-values: AUC difference vs Swedish reference (BH adjusted)")

# SES vs High
ref <- subset(data_original, ses == "High")
roc_ref <- roc(ref$cancer, ref$ai_score, quiet = TRUE)

groups_ses <- setdiff(levels(data_original$ses), "High")
res_delong_ses <- tibble(ses = character(), p_value = numeric())

for(gname in groups_ses){
  gdat <- subset(data_original, ses == gname)
  if(length(unique(gdat$cancer)) < 2) next
  roc_g <- roc(gdat$cancer, gdat$ai_score, quiet = TRUE)
  p <- roc.test(roc_g, roc_ref, method = "delong", paired = FALSE)$p.value
  res_delong_ses <- bind_rows(res_delong_ses, tibble(ses = gname, p_value = p))
}

res_delong_ses <- res_delong_ses %>% mutate(p_adj = p.adjust(p_value, method = "BH"))

kable(res_delong_ses, digits = 4,
      caption = "DeLong test p-values: AUC difference vs High SES reference (BH adjusted)")
```

# 4 Calibration

## 4.1 Overall calibration

```{r calibration_overall}

n_bins <- 10

data_original$bin <- cut(
  data_original$ai_score,
  breaks = seq(0, 1, length.out = n_bins + 1),
  include.lowest = TRUE
)

cal_overall <- data_original %>%
  group_by(bin) %>%
  summarise(n = n(), obs = mean(cancer), pred = mean(ai_score), .groups = "drop")

brier_overall <- mean((data_original$ai_score - data_original$cancer)^2)

ece_overall <- cal_overall %>%
  mutate(w = n/sum(n), gap = abs(obs - pred)) %>%
  summarise(ece = sum(w*gap)) %>%
  pull(ece)

kable(cal_overall, digits=4, caption="Overall calibration table (10 equal-width bins)")
kable(tibble(Brier = brier_overall, ECE_10bins = ece_overall), digits=4,
      caption="Overall calibration summary")

ggplot(cal_overall, aes(x = pred, y = obs)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line(linewidth=1) +
  geom_point(size=2) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  labs(
    title = "Calibration curve (overall)",
    subtitle = paste0(
      "Brier = ", round(brier_overall, 4),
      ", ECE(10 bins) = ", round(ece_overall, 4)
    ),
    x = "Mean predicted risk (ai_score)",
    y = "Observed cancer rate"
  ) +
  theme_classic(base_size = 12)
```

Interpretation: Despite very strong discrimination, calibration is not perfect (non-trivial ECE). This is consistent with a score that is primarily optimized for ranking.

## 4.2 Calibration by subgroup

```{r calibration_subgroups}
cal_eth <- data_original %>%
  group_by(ethnicity, bin) %>%
  summarise(n=n(), obs=mean(cancer), pred=mean(ai_score), .groups="drop")

ggplot(cal_eth, aes(pred, obs)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line() +
  geom_point(size=1.5) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  facet_wrap(~ ethnicity) +
  labs(title = "Calibration curves by ethnicity",
       x = "Mean predicted risk", y = "Observed rate") +
  theme_classic(base_size = 11)

brier_eth <- data_original %>%
  group_by(ethnicity) %>%
  summarise(
    n = n(),
    cases = sum(cancer),
    brier = mean((ai_score - cancer)^2),
    .groups = "drop"
  ) %>% arrange(desc(brier))

ece_eth <- cal_eth %>%
  group_by(ethnicity) %>%
  summarise(
    ece = sum((n/sum(n)) * abs(obs - pred)),
    .groups = "drop"
  ) %>% arrange(desc(ece))

kable(brier_eth, digits=4, caption="Brier score by ethnicity")
kable(ece_eth, digits=4, caption="ECE by ethnicity (10 bins)")

cal_ses <- data_original %>%
  group_by(ses, bin) %>%
  summarise(n=n(), obs=mean(cancer), pred=mean(ai_score), .groups="drop")

ggplot(cal_ses, aes(pred, obs)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line() +
  geom_point(size=1.5) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  facet_wrap(~ ses) +
  labs(title = "Calibration curves by SES",
       x = "Mean predicted risk", y = "Observed rate") +
  theme_classic(base_size = 11)

brier_ses <- data_original %>%
  group_by(ses) %>%
  summarise(
    n = n(),
    cases = sum(cancer),
    brier = mean((ai_score - cancer)^2),
    .groups = "drop"
  ) %>% arrange(desc(brier))

ece_ses <- cal_ses %>%
  group_by(ses) %>%
  summarise(
    ece = sum((n/sum(n)) * abs(obs - pred)),
    .groups = "drop"
  ) %>% arrange(desc(ece))

kable(brier_ses, digits=4, caption="Brier score by SES")
kable(ece_ses, digits=4, caption="ECE by SES (10 bins)")
```

# 5 Confounding and adjustment

## 5.1 Multivariable adjustment models

```{r adjustment_models}
data_original <- data_original %>%
  mutate(
    ethnicity = relevel(ethnicity, ref = "Swedish"),
    ses = relevel(ses, ref = "High")
  )

m0 <- glm(cancer ~ ai_score + ethnicity + ses,
          data = data_original, family = binomial())

m1 <- glm(cancer ~ ai_score + ethnicity + ses + region + vendor,
          data = data_original, family = binomial())

m2 <- glm(cancer ~ ai_score + ethnicity + ses + region + vendor + age + density,
          data = data_original, family = binomial())

tidy_or <- function(fit, label){
  broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE) %>%
    mutate(model = label) %>%
    select(model, term, estimate, conf.low, conf.high, p.value)
}

tab <- bind_rows(
  tidy_or(m0, "M0: + ethnicity + SES"),
  tidy_or(m1, "M1: + region + vendor"),
  tidy_or(m2, "M2: + age + density")
) %>%
  filter(
    term %in% c("ai_score",
               "ethnicityAfrican","ethnicityAsian","ethnicityOther","ethnicityOther European",
               "sesLow","sesMedium")
  ) %>%
  mutate(
    OR_CI = paste0(round(estimate, 3), " (", round(conf.low, 3), "–", round(conf.high, 3), ")"),
    p = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  ) %>%
  select(model, term, OR_CI, p)

kable(tab, caption = "Adjusted odds ratios (OR, 95% CI) across models")
```

# 6 Key findings, limitations, and next steps

## 6.1 Key findings

- **Discrimination:** Overall AUC is very high; subgroup AUCs are similarly high with small absolute differences.
- **Calibration:** Overall miscalibration is present (ECE), with some subgroup variation in Brier/ECE.
- **Adjustment:** After accounting for region/vendor (and age/density), the AI score remains strongly predictive; demographic terms can indicate group shifts at a fixed score (consistent with subgroup miscalibration or simulation-imposed shifts).

## 6.2 Practical relevance

When discrimination is near-perfect, **calibration and threshold-based metrics** become more informative for operational decisions (e.g., recall thresholds). A model can rank well (high AUC) but still over-/under-estimate risk in specific groups.

## 6.3 Limitations

- Some minority and intersectional subgroups are small; intersectional comparisons are less stable.
- Calibration metrics depend on binning; ECE can change with the binning strategy.
- The data are simulated; results illustrate an evaluation workflow rather than real-world performance.
