---
title: "equity-evaluation-mammo-ai"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

## Background and objectives
### Background

* This project evaluates whether a mammography AI algorithm’s cancer detection score performs consistently across demographic subgroups. Using a simulated screening dataset (N = 5,000), we quantify:

* Discrimination: how well the AI score separates cancer vs non-cancer (AUC-ROC).

* Calibration: whether the score behaves like a probability (predicted risk matches observed cancer rates).

* Potential confounding/structure: whether region- and vendor-related differences may influence apparent subgroup differences.

* Note: In many real systems, a model output can be a suspicion score rather than a calibrated probability. In that case, high AUC can coexist with poor calibration.

### Objectives

* Estimate overall and subgroup AUC-ROC (ethnicity, SES, and ethnicity × SES).

* Compare subgroup AUCs to reference groups (Swedish; High SES) using DeLong tests.

* Assess overall and subgroup calibration using calibration curves, Brier score, and ECE.

* Explore vendor/region structure and examine whether associations persist after adjustment (region/vendor/age/density).

## Data & preprocessing
```{r packages}
# Fix: set CRAN mirror so install.packages() works during knitting
options(repos = c(CRAN = "https://cloud.r-project.org"))

pkgs <- c("dplyr","knitr","pROC","ggplot2","scales","broom","tidyr","forcats")

need <- pkgs[!pkgs %in% rownames(installed.packages())]
if(length(need) > 0) install.packages(need)

invisible(lapply(pkgs, library, character.only = TRUE))

```

```{r}
data_path <- "C:/Users/22too/OneDrive/Desktop/r_task/mammography_ai_equity_data_simulated.csv"
data_original <- read.csv(data_path, stringsAsFactors = FALSE)

# Type casting + derived intersectional subgroup
data_original <- data_original %>% 
  mutate(
    cancer = as.integer(cancer),
    ethnicity = as.factor(ethnicity),
    ses = as.factor(ses),
    region = as.factor(region),
    vendor = as.factor(mammography_vendor),     # A/B/C/D only
    density = as.factor(breast_density),
    eth_ses = interaction(ethnicity, ses, drop = TRUE, sep = "/")
  ) %>% 
  select(-mammography_vendor, -breast_density)

head(data_original)

```

```{r}
miss <- data_original %>% summarise(across(everything(), ~sum(is.na(.))))
kable(miss, caption="Missing values per variable")
```

Preprocessing summary

* Cast categorical variables as factors and ensured cancer is 0/1 integer.

* Created eth_ses (ethnicity × SES) for intersectional summaries.

* Confirmed no missingness in the analytic variables.

### Subgroup sizes
```{r subgroup_sizes}
by_eth <- data_original %>% 
  group_by(ethnicity) %>% 
  summarise(n=n(), cases=sum(cancer), prev=mean(cancer), .groups="drop")

by_ses <- data_original %>% 
  group_by(ses) %>% 
  summarise(n=n(), cases=sum(cancer), prev=mean(cancer), .groups="drop")

by_int <- data_original %>% 
  group_by(eth_ses) %>% 
  summarise(n=n(), cases=sum(cancer), prev=mean(cancer), .groups="drop")

kable(by_eth, digits=3, caption="Group size by ethnicity")
kable(by_ses, digits=3, caption="Group size by SES")
kable(by_int, digits=3, caption="Group size by ethnicity × SES")

```

## Discrimination (AUC-ROC)
### Overall AUC
```{r roc_overall}
roc_all <- roc(response = data_original$cancer, predictor = data_original$ai_score, quiet=TRUE)
auc_all <- auc(roc_all)
ci_all <- ci.auc(roc_all, method="delong")

auc_all
ci_all

```

```{r}
roc_df <- data.frame(
  fpr = 1 - roc_all$specificities,
  tpr = roc_all$sensitivities
)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line(linewidth=1) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  labs(
    title = "Overall ROC curve",
    subtitle = paste0("AUC = ", round(as.numeric(auc_all), 3),
                      " (95% CI ", round(as.numeric(ci_all)[1],3), "–", round(as.numeric(ci_all)[3],3), ")"),
    x = "False positive rate (1 - specificity)",
    y = "True positive rate (sensitivity)"
  ) +
  theme_classic(base_size = 12)

```

Result (overall): Discrimination is extremely high (AUC close to 1).
This implies the score ranks cancer cases above non-cases very consistently in this simulated dataset.

### Subgroup AUCs
```{r auc_stratified}
auc_ethnicity <- data_original %>% 
  group_by(ethnicity) %>% 
  summarise(
    n = n(),
    cases = sum(cancer),
    auc = as.numeric(auc(roc(cancer, ai_score, quiet = TRUE))),
    ci_low = as.numeric(ci.auc(roc(cancer, ai_score, quiet = TRUE), method = "delong"))[1],
    ci_high = as.numeric(ci.auc(roc(cancer, ai_score, quiet = TRUE), method = "delong"))[3],
    .groups = "drop"
  )

kable(auc_ethnicity, digits=3, caption="AUC by ethnicity (95% CI: DeLong)")
```

```{r}
auc_ses <- data_original %>% 
  group_by(ses) %>% 
  summarise(
    n = n(),
    cases = sum(cancer),
    auc = as.numeric(auc(roc(cancer, ai_score, quiet = TRUE))),
    ci_low = as.numeric(ci.auc(roc(cancer, ai_score, quiet = TRUE), method = "delong"))[1],
    ci_high = as.numeric(ci.auc(roc(cancer, ai_score, quiet = TRUE), method = "delong"))[3],
    .groups = "drop"
  )

kable(auc_ses, digits=3, caption="AUC by SES (95% CI: DeLong)")
```

```{r}
min_cases <- 20

auc_int <- data_original %>%
  group_by(ethnicity, ses) %>%
  filter(sum(cancer) >= min_cases, sum(1 - cancer) >= min_cases) %>%
  summarise(
    n = n(),
    cases = sum(cancer),
    auc = as.numeric(auc(roc(cancer, ai_score, quiet = TRUE))),
    ci_low  = as.numeric(ci.auc(roc(cancer, ai_score, quiet = TRUE), method="delong"))[1],
    ci_high = as.numeric(ci.auc(roc(cancer, ai_score, quiet = TRUE), method="delong"))[3],
    .groups = "drop"
  )

kable(auc_int, digits=3,
      caption = paste0("AUC by ethnicity × SES (cells with ≥", min_cases, " cases and non-cases)"))

```

```{r}
auc_eth_plot <- auc_ethnicity %>% mutate(ethnicity = reorder(ethnicity, auc))

ggplot(auc_eth_plot, aes(x = auc, y = ethnicity)) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.2) +
  geom_point(size = 2.3) +
  geom_vline(xintercept = as.numeric(auc_all), linetype="dashed") +
  coord_cartesian(xlim = c(0.9, 1)) +
  labs(
    title = "Discrimination by ethnicity",
    subtitle = "Points: AUC; bars: 95% CI (DeLong). Dashed line: overall AUC.",
    x = "AUC-ROC",
    y = NULL
  ) +
  theme_classic(base_size = 12)

auc_ses_plot <- auc_ses %>% mutate(ses = reorder(ses, auc))

ggplot(auc_ses_plot, aes(x = auc, y = ses)) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.2) +
  geom_point(size = 2.3) +
  geom_vline(xintercept = as.numeric(auc_all), linetype="dashed") +
  coord_cartesian(xlim = c(0.9, 1)) +
  labs(
    title = "Discrimination by socioeconomic status (SES)",
    subtitle = "Points: AUC; bars: 95% CI (DeLong). Dashed line: overall AUC.",
    x = "AUC-ROC",
    y = NULL
  ) +
  theme_classic(base_size = 12)

```

Interpretation:
* Subgroup AUCs are all near the ceiling. In such settings, even if small differences exist, they may have limited practical meaning and are hard to distinguish reliably without very large samples or threshold-based evaluation.

### AUC comparisons vs reference groups
```{r delong_vs_swedish}
sw <- subset(data_original, ethnicity == "Swedish")
roc_sw <- roc(sw$cancer, sw$ai_score, quiet = TRUE)

groups_eth <- setdiff(levels(data_original$ethnicity), "Swedish")

res_delong <- data.frame(ethnicity = character(), p_value = numeric())

for(gname in groups_eth){
  gdat <- subset(data_original, ethnicity == gname)
  if(length(unique(gdat$cancer)) < 2) next
  roc_g <- roc(gdat$cancer, gdat$ai_score, quiet = TRUE)
  p <- roc.test(roc_g, roc_sw, method = "delong", paired = FALSE)$p.value
  res_delong <- rbind(res_delong, data.frame(ethnicity = gname, p_value = p))
}

kable(res_delong, digits=4, caption="DeLong test p-values: AUC difference vs Swedish reference")

```

```{r}
ref <- subset(data_original, ses == "High")
roc_ref <- roc(ref$cancer, ref$ai_score, quiet = TRUE)

groups_ses <- setdiff(levels(data_original$ses), "High")

res_delong_ses <- data.frame(ses = character(), p_value = numeric())

for(gname in groups_ses){
  gdat <- subset(data_original, ses == gname)
  if(length(unique(gdat$cancer)) < 2) next
  roc_g <- roc(gdat$cancer, gdat$ai_score, quiet = TRUE)
  p <- roc.test(roc_g, roc_ref, method = "delong", paired = FALSE)$p.value
  res_delong_ses <- rbind(res_delong_ses, data.frame(ses = gname, p_value = p))
}

kable(res_delong_ses, digits=4, caption="DeLong test p-values: AUC difference vs High SES reference")

```


## Calibration
### Overall calibration
```{r calibration_overall}
n_bins <- 10
data_original$bin <- cut(
  data_original$ai_score,
  breaks = seq(0, 1, length.out = n_bins + 1),
  include.lowest = TRUE
)

cal_overall <- data_original %>% 
  group_by(bin) %>% 
  summarise(n = n(), obs = mean(cancer), pred = mean(ai_score), .groups = "drop")

brier_overall <- mean((data_original$ai_score - data_original$cancer)^2)

ece_overall <- cal_overall %>% 
  mutate(w = n/sum(n), gap = abs(obs - pred)) %>% 
  summarise(ece = sum(w*gap)) %>% 
  pull(ece)

kable(cal_overall, digits=4, caption="Overall calibration table (10 equal-width bins)")
kable(data.frame(Brier = brier_overall, ECE_10bins = ece_overall), digits=4,
      caption="Overall calibration summary")

```

```{r}
ggplot(cal_overall, aes(x = pred, y = obs)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line(linewidth=1) +
  geom_point(size=2) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  labs(
    title = "Calibration curve (overall)",
    subtitle = paste0("Brier = ", round(brier_overall, 4),
                      ", ECE(10 bins) = ", round(ece_overall, 4)),
    x = "Mean predicted risk (ai_score)",
    y = "Observed cancer rate"
  ) +
  theme_classic(base_size = 12)

```

Interpretation:
* Despite very strong discrimination, calibration is imperfect (non-trivial ECE). This is consistent with an output that behaves like a ranking/suspicion score rather than a well-calibrated probability.

### Calibration by subgroup
```{r calibration_by_group}
cal_eth <- data_original %>% 
  group_by(ethnicity, bin) %>% 
  summarise(n=n(), obs=mean(cancer), pred=mean(ai_score), .groups="drop") %>%
  arrange(ethnicity, bin)

ggplot(cal_eth, aes(pred, obs)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line() +
  geom_point(size=1.5) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  facet_wrap(~ ethnicity) +
  labs(title = "Calibration curves by ethnicity",
       x = "Mean predicted risk", y = "Observed rate") +
  theme_classic(base_size = 11)

```

```{r}
brier_eth <- data_original %>% 
  group_by(ethnicity) %>% 
  summarise(
    total_n = n(),
    cases = sum(cancer),
    brier = mean((ai_score - cancer)^2),
    .groups = "drop"
  ) %>% arrange(desc(brier))

ece_eth <- cal_eth %>% 
  group_by(ethnicity) %>% 
  summarise(
    n_total = sum(n),
    cases = sum(n*obs),
    ece = sum((n / sum(n))* abs(obs - pred)),
    .groups = "drop"
  ) %>% arrange(desc(ece))

kable(brier_eth, digits=4, caption="Brier score by ethnicity")
kable(ece_eth, digits=4, caption="ECE by ethnicity (10 bins)")

```

```{r}
cal_ses <- data_original %>% 
  group_by(ses, bin) %>% 
  summarise(n=n(), obs=mean(cancer), pred=mean(ai_score), .groups="drop") %>%
  arrange(ses, bin)

ggplot(cal_ses, aes(pred, obs)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line() +
  geom_point(size=1.5) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  facet_wrap(~ ses) +
  labs(title = "Calibration curves by SES",
       x = "Mean predicted risk", y = "Observed rate") +
  theme_classic(base_size = 11)

```

```{r}
brier_ses <- data_original %>% 
  group_by(ses) %>% 
  summarise(
    total_n = n(),
    cases = sum(cancer),
    brier = mean((ai_score - cancer)^2),
    .groups = "drop"
  ) %>% arrange(desc(brier))

ece_ses <- cal_ses %>% 
  group_by(ses) %>% 
  summarise(
    n_total = sum(n),
    cases = sum(n*obs),
    ece = sum((n / sum(n))* abs(obs - pred)),
    .groups = "drop"
  ) %>% arrange(desc(ece))

kable(brier_ses, digits=4, caption="Brier score by SES")
kable(ece_ses, digits=4, caption="ECE by SES (10 bins)")

```

```{r}
cal_int <- data_original %>%
  group_by(ethnicity, ses, bin) %>%
  summarise(n=n(), obs=mean(cancer), pred=mean(ai_score), .groups="drop") %>%
  filter(n > 0) %>%
  arrange(ethnicity, ses, bin)

ggplot(cal_int, aes(pred, obs, group=1)) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  geom_line(linewidth=0.6) +
  geom_point(size=1.0) +
  coord_equal(xlim=c(0,1), ylim=c(0,1)) +
  facet_grid(ethnicity ~ ses) +
  scale_x_continuous(breaks=c(0,0.5,1)) +
  scale_y_continuous(breaks=c(0,0.5,1)) +
  labs(title="Calibration curves by ethnicity × SES",
       x="Mean predicted risk", y="Observed rate") +
  theme_classic(base_size=11) +
  theme(axis.text.x = element_text(angle=45, hjust=1))

```

```{r}
brier_int <- data_original %>% 
  group_by(ethnicity, ses) %>% 
  summarise(
    total_n = n(),
    cases = sum(cancer),
    brier = mean((ai_score - cancer)^2),
    .groups="drop"
  ) %>% arrange(desc(brier))

ece_int <- cal_int %>%
  group_by(ethnicity, ses) %>%
  summarise(
    n_total = sum(n),
    cases = sum(n * obs),
    ece = sum((n/sum(n)) * abs(obs - pred)),
    .groups="drop"
  ) %>% arrange(desc(ece))

kable(brier_int, digits=4, caption="Brier score by ethnicity × SES")
kable(ece_int, digits=4, caption="ECE by ethnicity × SES (10 bins)")

```

Interpretation:
* Calibration is imperfect across all groups, with some variation in Brier/ECE. Because threshold decisions in screening depend on the score scale, calibration gaps can matter even when AUC is uniformly high.

## Confounding and adjustment
#### Structure checks: vendor/region and ethnicity/region

```{r}
tab_vr <- table(data_original$vendor, data_original$region)
tab_er <- table(data_original$ethnicity, data_original$region)

chisq_vr <- chisq.test(tab_vr)
chisq_er <- chisq.test(tab_er)

chisq_vr
chisq_er

```

```{r}
vr <- as.data.frame(prop.table(tab_vr, margin = 2))
colnames(vr) <- c("vendor", "region", "prop")

ggplot(vr, aes(x = region, y = vendor, fill = prop)) +
  geom_tile() +
  scale_fill_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Vendor distribution by region",
    subtitle = "Column percentages (within region)",
    x = "Region", y = "Vendor", fill = "%"
  ) +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

er <- as.data.frame(prop.table(tab_er, margin = 2))
colnames(er) <- c("ethnicity", "region", "prop")

ggplot(er, aes(x = region, y = ethnicity, fill = prop)) +
  geom_tile() +
  scale_fill_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Ethnicity distribution by region",
    subtitle = "Column percentages (within region)",
    x = "Region", y = "Ethnicity", fill = "%"
  ) +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Interpretation:

* Vendor varies strongly by region (χ² p-value typically very small), indicating a plausible source of technical heterogeneity.

* In this simulated dataset, ethnicity may or may not vary meaningfully by region; interpret the χ² result directly. Even if the test is not significant, region and vendor can still be included as adjustment variables for completeness.

### Multivariable adjustment models
```{r}
data_original <- data_original %>%
  mutate(
    ethnicity = relevel(ethnicity, ref = "Swedish"),
    ses = relevel(ses, ref = "High")
  )

m0 <- glm(cancer ~ ai_score + ethnicity + ses,
          data = data_original, family = binomial())

m1 <- glm(cancer ~ ai_score + ethnicity + ses + region + vendor,
          data = data_original, family = binomial())

m2 <- glm(cancer ~ ai_score + ethnicity + ses + region + vendor + age + density,
          data = data_original, family = binomial())

summary(m0)
summary(m1)
summary(m2)

```

Interpretation (adjustment models):

* ai_score remains a very strong predictor in all models (as expected given the high AUC).

* If ethnicity coefficients remain non-null after adjustment, that implies that at a fixed AI score the outcome probability differs by group—consistent with group-specific miscalibration (or simulation-imposed group shifts).

* SES terms may be weaker or non-significant after accounting for the AI score and other covariates (interpret based on the fitted results).

## Key findings, limitations, and next steps
### Key findings

* Discrimination: Very high overall AUC and similarly high subgroup AUCs.

* Calibration: Non-trivial miscalibration overall (ECE) and some subgroup variation in Brier/ECE.

* Structure: Vendor differs by region; adjustment models help assess whether demographic differences persist after accounting for region/vendor and other covariates.

### Practical relevance

When discrimination is near-perfect, calibration and threshold-based metrics become more informative for operational decisions (e.g., recall thresholds). A model can rank well (high AUC) but still assign scores that over-/under-estimate risk in specific groups.

### Limitations

* Some minority and intersectional subgroups are small; intersectional comparisons are less stable.

* Calibration metrics depend on binning; ECE can change with binning strategy.

* The data are simulated; results illustrate analytic workflow rather than real-world performance.
